---
title: "Thesis_RMD"
author: "Sarmistha Bhattacharyya"
date: "11/02/2024"
output:
  pdf_document: default
  html_document:
    df_print: paged
editor_options: 
  markdown: 
    wrap: 72
---

Load required libraries

```{r, warning=FALSE}
library(dplyr)
library(cluster)  # For k-means clustering
library(factoextra)  # For visualization
library(cluster)
library(factoextra)
library(fpc)
library(mclust)
library(ORIClust)           
library(ggplot2)
library(FNN)
library(dbscan)
library(opticskxi)
library(NbClust)
library(tidyr)

```

Data Loading and structuring Renaming the columns of the dataset
properly

```{r, warning=FALSE}
#Load the data to the R env

load("~/Thesis research/VPA.Gene.Without650.RData")

#rename the dataset
df <- ev.VPA.without650

#rename the columns of the dataset with concentration_rep# manner 
colnames(df) <- c("untr_rep1","untr_rep2","untr_rep3","untr_rep4","untr_rep5","untr_rep6","25mM_rep1","25mM_rep2","25mM_rep3","150mM_rep1","150mM_rep2","150mM_rep3","350mM_rep1","350mM_rep2","350mM_rep3","450mM_rep1","450mM_rep2","450mM_rep3","550mM_rep1","550mM_rep2","550mM_rep3","800mM_rep1","800mM_rep2","800mM_rep3","1000mM_rep1","1000mM_rep2","1000mM_rep3")

```

View the dataset

```{r, echo = FALSE}
 #$$$head(df)
```

Select top 1000 genes which has the highest variability within different
samples (or concentration)

```{r, warning=FALSE}

# 1. Calculate the variability for each gene across all concentrations
gene_variability <- apply(df, 1, var)

# 2. Combine gene names and variability into a data frame
gene_variability_df <- data.frame(gene = rownames(df), variability = gene_variability)

# 3. Sort the genes based on their variability
sorted_genes <- gene_variability_df %>% arrange(desc(variability))
####EXPERIMENT ON BOXPLOT STARTS
#selected_genes <- sorted_genes[901:1000, ]
#gene_names_to_select <- selected_genes$gene
#selected_genes_data <- df[gene_names_to_select, ]
####EXPERIMENT ON BOXPLOT ENDS

# 4. Select the top 1000 genes with the highest variability
top_1000_genes <- head(sorted_genes, 1000)

# Now we can use top_1000_genes to select the corresponding rows from your original dataset
gene_names_to_select <- top_1000_genes$gene
selected_genes_data <- df[gene_names_to_select, ]

```

creating the dataset where top 1000 most variabled genes are store

```{r, warning=FALSE}
gene_expression_data <- selected_genes_data[,]
```

Summary of the dataset

(for each samples)

```{r, warning=FALSE}
# Compute summary statistics for each column (sample)
summary_stats <- apply(gene_expression_data, 2, function(x) c(mean(x), median(x), sd(x), min(x), max(x)))

# Set column names to correspond to sample names
colnames(summary_stats) <- colnames(gene_expression_data)

# Set row names to correspond to the functions that is applied to
rownames(summary_stats) <- c("Mean","Median","SD","Min","Max")

# View the summary statistics with sample names
#$$$print(summary_stats)

```

Summary of the dataset

(for each genes)

```{r, warning=FALSE , echo = FALSE}

# Compute summary statistics
summary_stats <- apply(gene_expression_data, 1, function(x) c(mean(x), median(x), sd(x), min(x), max(x)))

# Set row names to correspond to gene names
rownames(summary_stats) <- c("Mean","Median","SD","Min","Max")

# View the summary statistics with gene names
#$$$print(summary_stats)

```

Exploratory Data Analysis (EDA):
#Boxplot

```{r, warning=FALSE}

# Assuming your dataset is loaded into a dataframe named selected_genes_data

# Convert the matrix to a dataframe
selected_genes_df <- as.data.frame(selected_genes_data)

# Reshape the dataframe from wide to long format
df_long <- gather(selected_genes_df, key = "Sample", value = "Expression")

# Define colors for each group of samples
sample_colors <- c(rep("blue", 6), rep("red", 3), rep("green", 3), rep("orange", 3), rep("purple", 3), rep("yellow", 3), rep("brown", 3), rep("pink", 3))

# Reorder the x-axis labels in the original order
df_long$Sample <- factor(df_long$Sample, levels = unique(df_long$Sample))

# Plot the data with custom colors and reordered x-axis labels
ggplot(df_long, aes(x = Sample, y = Expression, fill = Sample)) +
  geom_boxplot() +
  scale_fill_manual(values = sample_colors) +
  labs(title = "Gene Expressions Across Concentrations",
       x = "Sample", y = "Expression") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))  # Rotate x-axis labels

```
Matplot for 1000 genes 

```{r}
concentration_names <- c("untr_rep1","untr_rep2","untr_rep3","untr_rep4","untr_rep5","untr_rep6","25mM_rep1","25mM_rep2","25mM_rep3","150mM_rep1","150mM_rep2","150mM_rep3","350mM_rep1","350mM_rep2","350mM_rep3","450mM_rep1","450mM_rep2","450mM_rep3","550mM_rep1","550mM_rep2","550mM_rep3","800mM_rep1","800mM_rep2","800mM_rep3","1000mM_rep1","1000mM_rep2","1000mM_rep3")

# Assuming your data is stored in a matrix called "gene_data"
# gene_data should have 1000 rows (genes) and 14 columns (concentrations)
# Set smaller margins
par(mar = c(6.5, 3, 1, 7))  # c(bottom, left, top, right)
# Plot the data using matplot
matplot(t(gene_expression_data), type = "l", col = rainbow(100), xlab = " ", ylab = "Expression Level",xaxt = "n")
# Customize x-axis labels
axis(1, at = seq_along(concentration_names), labels = concentration_names, las = 2)

```

#Heatmap of gene expression data

samples vs. genes it tells which genes have similar pattern in their
outcomes at which samples

```{r, warning=FALSE}

# Adjust font size and width of the plot
heatmap(gene_expression_data, scale = "row", cexCol = 0.5, width = 20, main = "Heatmap of Gene Expression")


```

#Correlation Matrix and Heatmap of correlation matrix

to check which samples have similar genes behaviour

```{r, warning=FALSE}
# Compute correlation matrix
correlation_matrix <- cor(gene_expression_data)

##head(correlation_matrix)

# Visualize correlation matrix as a heatmap
heatmap(correlation_matrix, main = "Correlation Heatmap", cexCol = 0.5, width = 20, cexRow = 0.5, height = 20)

```

Application of clustering algorithms

#K-means algorithm

-- Step 1: Choosing the optimal value of k based on maximum silhoutte
coefficients Possible set of k is taken from [2, sqrt(n/2)] where n is
the number of data which is 1000 in our case.

```{r, warning=FALSE}

# Compute total within-cluster sum of squares (WCSS) for different values of k
wss <- sapply(2:24, function(k) {
  kmeans(gene_expression_data, centers = k)$tot.withinss
})

# Plot the elbow plot
plot(2:24, wss, type = "b", frame = FALSE, 
     xlab = "Number of clusters (k)", ylab = "Total within-cluster sum of squares (WCSS)",
     main = "Elbow Plot for Optimal k")
# Customize x-axis labels to show integers from 2 to 24
axis(1, at = 2:24)

```

So elbow is visible for k = 4

Step 2: Identify the optimal k value (that k value which is being
referred by most of the methods)

```{r, warning=FALSE}
# Use the elbow point as the optimal value of k
Result_kM <- NbClust(gene_expression_data, distance = "euclidean", method = "kmeans", min.nc = 2, max.nc = 24, index = "all")

```

Assign optimal k value to a variable from the earlier summary

```{r, warning=FALSE}

#optimal k value
optimal_kM <- 2

```

Step 4: Perform k means clustering #Using NBClust package

```{r, warning=FALSE}

# Perform k-means clustering with the optimal value of k
set.seed(1234)  # Set seed for reproducibility
kmeans_result <- kmeans(gene_expression_data, centers = optimal_kM, nstart = 10,
       algorithm = c("Hartigan-Wong", "Lloyd", "Forgy",
                     "MacQueen"), trace = FALSE)

# Get cluster assignments for each gene
cluster_assignments_kM <- kmeans_result$cluster

# Add cluster assignments to your dataset
gene_expression_data_with_clusters_kM <- cbind(gene_expression_data, cluster = cluster_assignments_kM)

# Visualize the clustering result
# If your data has more than 2 dimensions, you can use PCA for visualization
pca <- prcomp(gene_expression_data)
fviz_cluster(list(data = pca$x[,1:2], cluster = cluster_assignments_kM), geom = "point", stand = FALSE, palette = "jco", main = "PCA Plot with Clusters")

# Print cluster centers
#print(kmeans_result$centers)
# Print the k means cluster result
#print(kmeans_result$size)

```

MATPLOT FOR K MEANS WITH k = 2

```{r}

# Assuming you have already performed k-means clustering and have cluster assignments stored in cluster_assignments_kM

# Number of clusters
num_clusters <- max(cluster_assignments_kM)

# Create a new directory to save the plots
dir.create("cluster_plots")

# Loop through each cluster
for (i in 1:num_clusters) {
  # Subset the data for the current cluster
  cluster_data <- gene_expression_data[cluster_assignments_kM == i, ]
  
  # Set smaller margins
  #par(mar = c(6.5, 3, 1, 5))  # c(bottom, left, top, right)
  
  # Plot the data using matplot
  matplot(t(cluster_data), type = "l", col = rainbow(100), xlab = " ", ylab = "Expression Level", xaxt = "n")
  
  # Customize x-axis labels
  axis(1, at = seq_along(concentration_names), labels = concentration_names, las = 2)
  
  # Save the plot
  plot_filename <- paste0("cluster_plots/cluster_", i, "_matplot.png")
  dev.copy(png, plot_filename)
  dev.off()
}

```



#By using the k-means from the Elbow plot with k = 4

```{r}
optimal_kM_elbow <- 4
```

```{r}
set.seed(1234)  # Set seed for reproducibility
kmeans_result_elbow <- kmeans(gene_expression_data, centers = optimal_kM_elbow, nstart = 10)

# Get cluster assignments for each gene
cluster_assignments_kM_elbow <- kmeans_result_elbow$cluster

# Add cluster assignments to your dataset
gene_expression_data_with_clusters_kM_elbow <- cbind(gene_expression_data, cluster = cluster_assignments_kM_elbow)

#adding new code
# Calculate silhouette score
silhouette_score_elbow <- silhouette(cluster_assignments_kM_elbow, dist(gene_expression_data))

# Print silhouette score
#print(silhouette_score)

# Combine cluster assignments with silhouette scores
silhouette_df <- data.frame(Cluster = cluster_assignments_kM_elbow, Silhouette = silhouette_score_elbow[, "sil_width"])

# Calculate average silhouette score for each cluster
cluster_silhouette_elbow <- silhouette_df %>%
  group_by(Cluster) %>%
  summarize(Avg_Silhouette = mean(Silhouette))

# Print average silhouette score for each cluster
print(cluster_silhouette_elbow)

# Compute silhouette score for each data point
#silhouette_scores <- silhouette(cluster_assignments_kM, dist(gene_expression_data))

# Calculate total average silhouette score
total_avg_silhouette_elbow <- mean(silhouette_score_elbow[, "sil_width"])

# Print total average silhouette score
print(total_avg_silhouette_elbow)
#end of new code
# Visualize the clustering result
# If your data has more than 2 dimensions, you can use PCA for visualization
pca <- prcomp(gene_expression_data)
fviz_cluster(list(data = pca$x[,1:2], cluster = cluster_assignments_kM_elbow), geom = "point", stand = FALSE, palette = "jco", main = "PCA Plot with Clusters")

# Print cluster centers
#print(kmeans_result$centers)
# Print the k means cluster result
#print(kmeans_result$size)
# Calculate the size of each cluster
cluster_sizes_elbow <- table(cluster_assignments_kM_elbow)

# Print the size of each cluster
print(cluster_sizes_elbow)


```

MATPLOT OF EACH CLUSTERS FOR k = 4

```{r}

# Assuming you have already performed k-means clustering and have cluster assignments stored in cluster_assignments_kM

# Number of clusters
num_clusters <- max(cluster_assignments_kM_elbow)

# Create a new directory to save the plots
dir.create("cluster_plots")

# Loop through each cluster
for (i in 1:num_clusters) {
  # Subset the data for the current cluster
  cluster_data <- gene_expression_data[cluster_assignments_kM_elbow == i, ]
  
  # Set smaller margins
  #par(mar = c(6.5, 3, 1, 7))  # c(bottom, left, top, right)
  
  # Plot the data using matplot
  matplot(t(cluster_data), type = "l", col = rainbow(100), xlab = " ", ylab = "Expression Level", xaxt = "n")
  
  # Customize x-axis labels
  axis(1, at = seq_along(concentration_names), labels = concentration_names, las = 2)
  
  # Save the plot
  plot_filename <- paste0("cluster_plots/cluster_", i, "_matplot.png")
  dev.copy(png, plot_filename)
  dev.off()
}

```

Genes assigned on the clusters:

```{r warning=FALSE, echo = FALSE}

# Add cluster assignments to your dataset
gene_expression_data_with_clusters_kM <- cbind(gene_expression_data, cluster = cluster_assignments_kM)

# Create a data frame with gene names and their corresponding cluster assignments
gene_cluster_df_kM <- data.frame(Gene = rownames(gene_expression_data), Cluster = cluster_assignments_kM)

# Display gene names for each cluster
for (i in 1:optimal_kM) {
  cat("Genes in Cluster", i, ":\n")
  ####print(gene_cluster_df_kM$Gene[gene_cluster_df_kM$Cluster == i])
}

#Gene cluster 1  : "225846_at"    "228387_at"    "231899_at"    "205710_at"    "209398_at"   
# Gene Cluster 2: "210650_s_at"  "228708_at"    "219768_at"    "204949_at"    "229800_at"   

```

Step 5: Performance evaluation. cluster size + Silhoutte coeffiecient of
the final cluster (for NBClust package)

```{r, warning=FALSE, echo = FALSE}

#size of cluster
# Assuming nb_clust_result$Best.partition contains the cluster labels
cluster_labels_kM <- Result_kM$Best.partition

# Count the occurrences of each unique cluster label
cluster_sizes_kM <- table(cluster_labels_kM)

# Print the cluster sizes
print(cluster_sizes_kM)

# Extract Silhouette index values
silhouette_value_kM <- Result_kM$Best.nc["Value_Index", "Silhouette"]
#print("Silhoutte coefficient is :")
#silhouette_value_kM
cat("Silhouette coefficient is:", silhouette_value_kM, "\n")

```
Hirarchical clustering

--Step 1: Determine the optimal number of clusters using NbClustalue

```{r, warning=FALSE}

nb_clust_result_hc <- NbClust(gene_expression_data, distance = "euclidean", method = "ward.D2", min.nc = 2, max.nc = 24, index = "all")

```

Set the optimal k value to the variable to use it for further analysis

```{r, warning=FALSE}
optimal_kHC <- 4
```

--Step 2: Perform hierarchical clustering with the optimal number of
clusters

```{r, warning=FALSE}

hc <- hclust(dist(gene_expression_data), method = "ward.D2")
hc_clusters <- cutree(hc, k = optimal_kHC)

```

-- Step 3: Visualize the clustering result

```{r, warning=FALSE, echo = FALSE}

pca <- prcomp(gene_expression_data)
fviz_cluster(list(data = pca$x[,1:2], cluster = hc_clusters), geom = "point", stand = FALSE, palette = "jco", main = "PCA Plot with Clusters")

```

MATPLOT OF EACH CLUSTERS FOR hirarchical clustering for k = 4

```{r}

# Assuming you have already performed k-means clustering and have cluster assignments stored in cluster_assignments_kM

# Number of clusters
num_clusters <- max(hc_clusters)

# Create a new directory to save the plots
dir.create("cluster_plots")

# Loop through each cluster
for (i in 1:num_clusters) {
  # Subset the data for the current cluster
  cluster_data <- gene_expression_data[hc_clusters == i, ]
  
  # Set smaller margins
  #par(mar = c(6.5, 3, 1, 7))  # c(bottom, left, top, right)
  
  # Plot the data using matplot
  matplot(t(cluster_data), type = "l", col = rainbow(100), xlab = " ", ylab = "Expression Level", xaxt = "n")
  
  # Customize x-axis labels
  axis(1, at = seq_along(concentration_names), labels = concentration_names, las = 2)
  
  # Save the plot
  plot_filename <- paste0("cluster_plots/cluster_", i, "_matplot.png")
  dev.copy(png, plot_filename)
  dev.off()
}

```


-- Step 4: Print cluster centers (mean expression profiles)

```{r, warning=FALSE}

cluster_centers_HC <- aggregate(gene_expression_data, by = list(cluster = hc_clusters), FUN = mean)
#print(cluster_centers_HC)

```

--Step 5: Performance evaluation - Silhouette coefficient of the final
clusters

```{r, warning=FALSE, echo = FALSE}
#size of cluster
# Assuming nb_clust_result$Best.partition contains the cluster labels
cluster_labels_hc <- nb_clust_result_hc$Best.partition
# Count the occurrences of each unique cluster label
cluster_sizes_hc <- table(cluster_labels_hc)
# Print the cluster sizes
print(cluster_sizes_hc)

silhouette_value_hc <- nb_clust_result_hc$Best.nc["Value_Index", "Silhouette"]
#silhouette_value_hc
# Assuming silhouette_value_kM is the variable storing the silhouette coefficient value
cat("Silhouette coefficient for hirarchical clustering is:", silhouette_value_hc, "\n")

```

Silhoutte score for each clusters separately
 
```{r}

# Calculate silhouette score for each data point
silhouette_scores_hc <- silhouette(hc_clusters, dist(gene_expression_data))

# Aggregate silhouette scores by cluster
silhouette_df_hc <- data.frame(Cluster = hc_clusters, Silhouette = silhouette_scores_hc[, "sil_width"])
cluster_silhouette_hc <- silhouette_df_hc %>%
  group_by(Cluster) %>%
  summarize(Avg_Silhouette = mean(Silhouette))

# Print average silhouette score for each cluster
print(cluster_silhouette_hc)

```

#Print the genes with the clusters

```{r, warning=FALSE, echo = FALSE}

# Add cluster assignments to your dataset
gene_expression_data_with_clusters_hc <- cbind(gene_expression_data, cluster = cluster_labels_hc)

# Create a data frame with gene names and their corresponding cluster assignments
gene_cluster_df_hc <- data.frame(Gene = rownames(gene_expression_data), Cluster = cluster_labels_hc)

# Display gene names for each cluster
for (i in 1:optimal_kHC) {
  #cat("Genes in Cluster", i, ":\n")
  #print(gene_cluster_df_hc$Gene[gene_cluster_df_hc$Cluster == i])
}

#Cluster 1 : "202625_at"    "228708_at"    "204851_s_at"  "204949_at"    "229800_at"
#Cluster 2:  "210302_s_at"  "238878_at"    "219197_s_at"  "219557_s_at"  "222900_at" 
#Cluster 3: "227238_at"    "227241_at"    "219995_s_at"  "227856_at"    "227202_at"
#Cluster 4: "1554012_at"   "221086_s_at"  "242193_at"    "233972_s_at"  "233327_at"

```

DBSCAN

#DBSCAN Method 1 : Finding optimal epsilon using silhoutte coefficient

Step 1: Define calculate_silhouette() to calculte silhoutte coeffiencet
of each cluster

```{r}

# Function to calculate silhouette score for a given eps and MinPts
calculate_silhouette <- function(data, eps, MinPts) {
  dbscan_result <- dbscan(data, eps = eps, minPts = MinPts)
  if (length(unique(dbscan_result$cluster)) == 1) {
    return(NA)  # Skip if only noise points are detected
  }
  silhouette <- silhouette(dbscan_result$cluster, dist(data))
  return(mean(silhouette))
}

```

Step 2: Convert gene_expression_data to matrix

```{r}

gene_expression_matrix <- as.matrix(gene_expression_data)

```

Step 3: Parameter tuning of epsilon. Taking ranges of value between 0.1
to 10. set the parameters minPts = dim +1 = 27 + 1 = 28 (used 10) mhere

```{r, echo = FALSE}

# Grid search for epsilon and MinPts
eps_values <- seq(0.1, 10, by = 0.1)  # Example range for epsilon
MinPts_values <- 10  # Example range for MinPts
silhouette_scores <- matrix(NA, nrow = length(eps_values), ncol = length(MinPts_values))
for (i in seq_along(eps_values)) {
  for (j in seq_along(MinPts_values)) {
    silhouette_scores[i, j] <- calculate_silhouette(gene_expression_matrix, eps_values[i], MinPts_values[j])
  }
}

#print(silhouette_scores)
# Find optimal parameters
optimal_params <- which.max(silhouette_scores)
#print(optimal_params)
#print(optimal_params[1])
#print(eps_values)
#silhouette_scores[optimal_params]
optimal_eps <- eps_values[optimal_params[1]]

print(paste("Optimal epsilon based on silhoutte value is :", optimal_eps))

```

Step 4: DBSCAN algorithm

```{r, warning=FALSE, echo = FALSE}

# Perform DBSCAN clustering with optimal parameters
dbscan_result <- dbscan(gene_expression_matrix, eps = optimal_eps, MinPts = 10)

# Print number of clusters and cluster sizes
num_clusters_dbscan <- length(unique(dbscan_result$cluster)) - 1  # Subtract 1 for noise points
cluster_sizes_dbscan <- table(dbscan_result$cluster)
print(paste("Optimal number of clusters:", num_clusters_dbscan))
print("Cluster sizes:")
print(cluster_sizes_dbscan)

```

Step 5: Final cluster result

```{r, echo = FALSE}
# Extract cluster labels
cluster_labels_dbscan <- dbscan_result$cluster
# Calculate silhouette score
silhouette_avg_dbscan <- silhouette(cluster_labels_dbscan, dist(gene_expression_matrix))
# Print the silhouette score
print(paste("Average silhouette score:", mean(silhouette_avg_dbscan[, "sil_width"])))
# Extract cluster labels
cluster_labels_dbscan <- dbscan_result$cluster
# Calculate silhouette score
silhouette_avg_dbscan <- silhouette(cluster_labels_dbscan, dist(gene_expression_matrix))
# Combine cluster labels and silhouette scores into a data frame
silhouette_df_dbscan <- data.frame(Cluster = cluster_labels_dbscan, Silhouette = silhouette_avg_dbscan[, "sil_width"])
# Calculate average silhouette score for each cluster
cluster_silhouette_dbscan <- silhouette_df_dbscan %>%
  group_by(Cluster) %>%
  summarize(Avg_Silhouette = mean(Silhouette))
# Print cluster sizes and average silhouette score for each cluster
print(cluster_silhouette_dbscan)

```

MATPLOT OF EACH CLUSTERS FOR DBSCAN for EPSILON = 2.6

```{r}

# Assuming you have already performed k-means clustering and have cluster assignments stored in cluster_assignments_kM

# Number of clusters
num_clusters <- max(cluster_labels_dbscan)

# Create a new directory to save the plots
dir.create("cluster_plots")

# Loop through each cluster
for (i in 1:num_clusters) {
  # Subset the data for the current cluster
  cluster_data <- gene_expression_data[cluster_labels_dbscan == i, ]
  
  # Set smaller margins
  #par(mar = c(6.5, 3, 1, 7))  # c(bottom, left, top, right)
  
  # Plot the data using matplot
  matplot(t(cluster_data), type = "l", col = rainbow(100), xlab = " ", ylab = "Expression Level", xaxt = "n")
  
  # Customize x-axis labels
  axis(1, at = seq_along(concentration_names), labels = concentration_names, las = 2)
  
  # Save the plot
  plot_filename <- paste0("cluster_plots/cluster_", i, "_matplot.png")
  dev.copy(png, plot_filename)
  dev.off()
}

```


Step 6: Visualization using PCA

```{r, echo = FALSE}

# Perform PCA
pca_result <- prcomp(gene_expression_data, scale = TRUE)

# Visualize PCA plot with cluster assignments
fviz_pca_ind(pca_result, geom = "point", habillage = as.factor(dbscan_result$cluster), 
             title = "PCA plot of DBSCAN Clustering Results")

```

#show the individual genes assigned to the clusters

```{r, echo = FALSE}
# Create a data frame with gene names and their corresponding cluster assignments
gene_cluster_df_dbscan <- data.frame(Gene = rownames(gene_expression_data), Cluster = cluster_labels_dbscan)

# Display gene names for each cluster
for (i in 1:length(unique(cluster_labels_dbscan))) {
  #cat("Genes in Cluster", i-1, ":\n")
  #print(gene_cluster_df_dbscan$Gene[gene_cluster_df_dbscan$Cluster == i-1])
}

#Gene cluster 0 (outliers): "239205_s_at"  "210302_s_at"  "238878_at"    "219557_s_at"  "230493_at" 

#Gene 1:"239206_at"    "205523_at"    "230204_at"    "227238_at"    "223044_at"   

#Gene 2: 210303_at"    "1556378_a_at" "237435_at"    "237322_at"    "204424_s_at" 


```

#DBSCAN Method 2 : Finding optimal epsilon using KNN Plot Optimal
epsilon value based on KNN plot

Step 1: set the parameters minPts = dim +1 = 27 + 1 = 28 here and find
the optimal epsilon value by analyzing the KNN distance plot.

```{r, warning=FALSE}

kNNdistplot(gene_expression_data, minPts = 10)
abline(h=4, col = "red", lty = 2)
# Noise seems to have start from 5.6

#setting the parameters value
epsion= 4#5.6
minPts= 10

```

It is evident from the KNN plot that , the plot apperaed to have more
steap gradiant when epsilon is 4. So optimal value of epsilon seems to
be 7 from the KNN Plot.

Step2: DBSCAN algorithm

```{r, warning=FALSE}

res_dbscan <- dbscan(gene_expression_data, eps = epsion, minPts = minPts)
#$$$res_dbscan

# Check the border points
border_points <- which(res_dbscan$borderPoints)
#$$$print("Number of border points:")
#$$$print(border_points)
cluster_labels_dbscan <- res_dbscan$cluster

```

MATPLOT OF EACH CLUSTERS FOR DBSCAN for EPSILON = 4

```{r}

# Assuming you have already performed k-means clustering and have cluster assignments stored in cluster_assignments_kM

# Number of clusters
num_clusters <- max(cluster_labels_dbscan)

# Create a new directory to save the plots
dir.create("cluster_plots")

# Loop through each cluster
for (i in 1:num_clusters) {
  # Subset the data for the current cluster
  cluster_data <- gene_expression_data[cluster_labels_dbscan == i, ]
  
  # Set smaller margins
  #par(mar = c(6.5, 3, 1, 7))  # c(bottom, left, top, right)
  
  # Plot the data using matplot
  matplot(t(cluster_data), type = "l", col = rainbow(100), xlab = " ", ylab = "Expression Level", xaxt = "n")
  
  # Customize x-axis labels
  axis(1, at = seq_along(concentration_names), labels = concentration_names, las = 2)
  
  # Save the plot
  plot_filename <- paste0("cluster_plots/cluster_", i, "_matplot.png")
  dev.copy(png, plot_filename)
  dev.off()
}

```



So DBSCAN creates one cluster with 994 genes and 6 genes are declared as
outliers

Step3: Visualization of DBSCAN using PCA

```{r, warning=FALSE, echo = FALSE}
# Perform PCA
pca_result <- prcomp(gene_expression_data, scale = TRUE)

# Visualize PCA plot with cluster assignments
#$$$fviz_pca_ind(pca_result, geom = "point", habillage = as.factor(res_dbscan$cluster), 
#$$$             title = "PCA plot of DBSCAN Clustering Results")

```

-- Step 4: Performance evaluation of DBSCAN by Calculating Silhoutte
coefficient for each cluster

```{r, warning=FALSE}

# Compute the distance matrix
dist_matrix <- dist(gene_expression_data)

# Calculate the silhouette score for DBSCAN clustering result
cluster_labels_dbscan <- res_dbscan$cluster
silhouette_score_db <- silhouette(cluster_labels_dbscan, dist_matrix)

# Print the silhouette score
summary(silhouette_score_db)

# Calculate the total silhouette score
total_silhouette_score <- mean(silhouette_score_db[, "sil_width"])
print(paste("Total silhouette score:", total_silhouette_score))


```

```{r, echo = FALSE}

# Create a data frame with gene names and their corresponding cluster assignments
gene_cluster_df_dbscan <- data.frame(Gene = rownames(gene_expression_data), Cluster = cluster_labels_dbscan)

# Display gene names for each cluster
for (i in 1:length(unique(cluster_labels_dbscan))) {
  #cat("Genes in Cluster", i-1, ":\n")
  #print(gene_cluster_df_dbscan$Gene[gene_cluster_df_dbscan$Cluster == i-1])
}

```

OPTICS

1.  Step 1: Application of OPTICS using same minPts value. lower limit
    of eps_cl is set as 4.6 which is adoped from the optimal epsilon
    value for silhoutte score from previously

```{r, warning=FALSE}
# Compute the distance matrix
dist_matrix <- dist(gene_expression_data)

res_optics <- optics(dist_matrix, minPts = 10)
#res_optics
#plot(res_optics)
res_optics <- extractDBSCAN(res_optics, eps_cl = 4)
res_optics
#hullplot(gene_expression_data, res_optics)

pca <- prcomp(gene_expression_data)

fviz_cluster(list(data = pca$x[,1:2], cluster = res_optics$cluster), geom = "point", stand = FALSE, palette = "jco", main = "PCA Plot with Clusters")



```

--Step 2: Performance alanysis of the OPTICS cluster by Calculation of
Silhoutte coefficients

```{r, warning=FALSE}

# Compute the distance matrix
dist_matrix <- dist(gene_expression_data)

# Calculate the silhouette score for DBSCAN clustering result
cluster_labels_optics <- res_optics$cluster
silhouette_score_optics <- silhouette(cluster_labels_optics, dist_matrix)

# Print the silhouette score
#summary(silhouette_score_optics)

cat("Silhouette coefficient for OPTICS clustering is:",mean(silhouette_score_optics[, "sil_width"]), "\n")

```


MATPLOT OF EACH CLUSTERS FOR OPTICS 

```{r}

# Assuming you have already performed k-means clustering and have cluster assignments stored in cluster_assignments_kM

# Number of clusters
num_clusters <- max(cluster_labels_optics)

# Create a new directory to save the plots
dir.create("cluster_plots")

# Loop through each cluster
for (i in 1:num_clusters) {
  # Subset the data for the current cluster
  cluster_data <- gene_expression_data[cluster_labels_optics == i, ]
  
  # Set smaller margins
  #par(mar = c(6.5, 3, 1, 7))  # c(bottom, left, top, right)
  
  # Plot the data using matplot
  matplot(t(cluster_data), type = "l", col = rainbow(100), xlab = " ", ylab = "Expression Level", xaxt = "n")
  
  # Customize x-axis labels
  axis(1, at = seq_along(concentration_names), labels = concentration_names, las = 2)
  
  # Save the plot
  plot_filename <- paste0("cluster_plots/cluster_", i, "_matplot.png")
  dev.copy(png, plot_filename)
  dev.off()
}

```


# display the name of each genes and the clusters

```{r}
# Create a data frame with gene names and their corresponding cluster assignments
gene_cluster_df_optics <- data.frame(Gene = rownames(gene_expression_data), Cluster = cluster_labels_optics)

# Display gene names for each cluster
for (i in 1:length(unique(cluster_labels_optics))) {
  #cat("Genes in Cluster", i-1, ":\n")
  #print(gene_cluster_df_optics$Gene[gene_cluster_df_optics$Cluster == i-1])
}

# Gene 0 (Outliers):"239205_s_at"  "210302_s_at"  "205523_at"    "230204_at"    "1554012_at"  
#Gene 1: "243278_at"    "229724_at"    "226731_at"    "232327_at"    "203304_at"   
#Gene 2 : "228387_at"    "231899_at"    "230198_at"    "202708_s_at"  "205710_at" 

```

Silhoutte score of each cluster

```{r}

# Calculate the silhouette score for each data point
silhouette_scores_optics <- silhouette(cluster_labels_optics, dist_matrix)

# Combine cluster assignments with silhouette scores
silhouette_df_optics <- data.frame(Cluster = cluster_labels_optics, Silhouette = silhouette_scores_optics[, "sil_width"])

# Calculate average silhouette score for each cluster
cluster_silhouette_optics <- silhouette_df_optics %>%
  group_by(Cluster) %>%
  summarize(Avg_Silhouette = mean(Silhouette))

# Print average silhouette score for each cluster
print(cluster_silhouette_optics)

```

Several methods to extract a clustering from the order returned by
OPTICS are implemented:

extractDBSCAN() extracts a clustering from an OPTICS ordering that is
similar to what DBSCAN would produce with an eps set to eps_cl (see
Ankerst et al, 1999). The only difference to a DBSCAN clustering is that
OPTICS is not able to assign some border points and reports them instead
as noise. #we used extractDBSCAN() extractXi() extract clusters
hierarchically specified in Ankerst et al (1999) based on the steepness
of the reachability plot. One interpretation of the xi parameter is that
it classifies clusters by change in relative cluster density. The used
algorithm was originally contributed by the ELKI framework and is
explained in Schubert et al (2018), but contains a set of fixes.

ORICC

```{r, warning=FALSE}

col_ID <- c(1:1000)
gene_expression_data <- cbind(gene_expression_data,col_ID)

oric.clus<-ORICC1(gene_expression_data, data.col= 1:27, id.col = 28 , n.rep= c(6,rep(3,7)),plot.format = "jpg" , name.profile = "all", onefile = FALSE )

oric_cluster<-oric.clus$cluster

table(oric.clus$cluster)

barplot(table(oric.clus$cluster), las = 1)

```

Plot of the clusters using different clustering algorithm using PCA 1
and PCA 2

```{r}

# Extract the percentage of variance explained by each principal component
variance_explained <- round(summary(pca)$importance[2, 1:2] * 100, 2)

fviz_cluster(list(data = pca$x[,1:2], cluster = cluster_assignments_kM), geom = "point", stand = FALSE, palette = "jco", main = paste("PCA Plot with Clusters\n", "PC1:", variance_explained[1], "%, PC2:", variance_explained[2], "%"))

fviz_cluster(list(data = pca$x[,1:2], cluster = hc_clusters), geom = "point", stand = FALSE, palette = "jco", main = paste("PCA Plot with Clusters\n", "PC1:", variance_explained[1], "%, PC2:", variance_explained[2], "%"))

# Visualize PCA plot with cluster assignments
#fviz_pca_ind(pca_result, geom = "point", habillage = as.factor(dbscan_result$cluster), 
#             title = "PCA plot of DBSCAN Clustering algorithm")

fviz_cluster(list(data = pca$x[,1:2], cluster = dbscan_result$cluster), geom = "point", stand = FALSE, palette = "jco", main = paste("PCA Plot with Clusters\n", "PC1:", variance_explained[1], "%, PC2:", variance_explained[2], "%"))

fviz_cluster(list(data = pca$x[,1:2], cluster = res_optics$cluster), geom = "point", stand = FALSE, palette = "jco", main = paste("PCA Plot with Clusters\n", "PC1:", variance_explained[1], "%, PC2:", variance_explained[2], "%"))

```

Cluster size and silhoutte score of each clusters and average silhoutte
score of the algorithm

```{r, warning=FALSE}
# Assuming nb_clust_result$Best.partition contains the cluster labels
cluster_labels_kM <- Result_kM$Best.partition
# Count the occurrences of each unique cluster label
cluster_sizes_kM <- table(cluster_labels_kM)
# Print the cluster sizes
print(cluster_sizes_kM)
# Extract Silhouette index values
silhouette_value_kM <- Result_kM$Best.nc["Value_Index", "Silhouette"]
cat("Silhouette coefficient for k-means clustering algorithm is:", silhouette_value_kM, "\n")

# Assuming nb_clust_result$Best.partition contains the cluster labels
cluster_labels_hc <- nb_clust_result_hc$Best.partition
# Count the occurrences of each unique cluster label
cluster_sizes_hc <- table(cluster_labels_hc)
# Print the cluster sizes
print(cluster_sizes_hc)
silhouette_value_hc <- nb_clust_result_hc$Best.nc["Value_Index", "Silhouette"]
# Assuming silhouette_value_kM is the variable storing the silhouette coefficient value
cat("Silhouette coefficient for hirarchical clustering is:", silhouette_value_hc, "\n")

# Perform DBSCAN clustering with optimal parameters
dbscan_result <- dbscan(gene_expression_matrix, eps = optimal_eps, MinPts = 28)
# Print number of clusters and cluster sizes
num_clusters_dbscan <- length(unique(dbscan_result$cluster)) - 1  # Subtract 1 for noise points
cluster_sizes_dbscan <- table(dbscan_result$cluster)
#print(paste("Optimal number of clusters:", num_clusters_dbscan))
print("DBSCAN Cluster sizes:")
print(cluster_sizes_dbscan)
cat("Silhouette coefficient for DBSCAN clustering is:", mean(silhouette_avg_dbscan[, "sil_width"]), "\n")

cluster_sizes_optics <- table(res_optics$cluster)
#print(paste("Optimal number of clusters:", num_clusters_dbscan))
print("Optics Cluster sizes:")
print(cluster_sizes_optics)
cat("Silhouette coefficient for OPTICS clustering is:", mean(silhouette_score_optics[, "sil_width"]), "\n")

```

```{r}
library(cluster)


oric.clus<-ORICC1(gene_expression_data, data.col= 1:27, id.col = 28 , n.rep= c(6,rep(3,7)) , plot.format = "jpg" , name.profile = "all", onefile = FALSE )

table(oric.clus$cluster)

barplot(table(oric.clus$cluster), las = 1)

original_genes <- 1:1000

# Selected genes by ORICC
selected_genes <- unique(oric.clus$top.id)

# Genes eliminated by ORICC
eliminated_genes <- setdiff(original_genes, selected_genes)

# Print the eliminated genes
print(eliminated_genes)

# Define the indices of rows to be removed
rows_to_remove <- eliminated_genes

# Define the indices of rows to be removed
#rows_to_remove <- c(278, 391, 551, 615, 622, 626, 756, 879, 918, 935, 938, 985)
# Remove the specified rows from the gene_expression_data matrix
gene_expression_data_filtered <- gene_expression_data[-rows_to_remove, ]
# Assuming oric.clus$cluster contains the cluster assignments
# and oric.clus$cluster is a vector of cluster assignments for each data point

# Compute silhouette
sil <- silhouette(oric.clus$cluster, dist(gene_expression_data_filtered))

# Mean silhouette width
mean_sil_width <- mean(sil[, "sil_width"])

mean_sil_width
library(factoextra)

# Assuming oric.clus$cluster contains the cluster assignments
# and oric.clus$cluster is a vector of cluster assignments for each data point
pca <- prcomp(gene_expression_data_filtered)
# Visualize clusters
#fviz_cluster(list(data = gene_expression_data_filtered, cluster = oric.clus$cluster))

# Visualize clusters using PCA with automatic coloring
fviz_cluster(list(data = pca$x[,1:2], cluster = oric.clus$cluster), geom = "point", stand = FALSE, main = "PCA Plot with Clusters for ORICC", palette = "default")



```

Note that the `echo = FALSE` parameter was added to the code chunk to
prevent printing of the R code that generated the plot.
